### 2023-04-21 doc/response.tex

\usepackage{lineno}
\linenumbers

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript.
We have tried to highlight the limitation of our correlation analysis and avoid linking its results to any causal statements.
We have also tried to clarify that the causal conclusions are supported by the modeling analysis, rather than the correlation analysis.
Below please find our detailed responses to reviewers.

@@

\revtext{Reviewer \#1 Comments for the Author:}

\revtext{The authors have done a good job responding to my previous set of comments and have made appropriate revisions to their manuscript.}

@@

Thank you for your review.

\revtext{Examining the updated timeseries results, my remaining concern regards the use of Pearson correlation of weekly case incidence in the analysis of dynamic coupling between community and campus, which features strongly in the significance statement and generally as a contribution of the paper.}

\revtext{I'm not sure it's appropriate to give p values based on a null model of independent identically distributed samples, when timeseries data like this is autocorrelated and non-stationary. If possible this should be reviewed by someone with a background in analysis of coupled timeseries data. At minimum, the methods used to compute correlation should be given in more detail and the text should be examined to ensure that causal inference is not being overstated.}

Thank you for pointing this out.
We recognize that the causal language has been too strong.
In particular, the causal inference is primarily supported by the mathematical modeling analysis, rather than the correlation analysis.
We have revised the paper throughout to make the distinction clear and not overstate the implications of the correlations.
Overall, we have tried to avoid linking correlations to causal statements.

We have also added the following paragraph in the beginning of the results section to make the limitation of the method clearer:

``
In order to compare the transmission dynamics on PU campus and in nearby communities, we calculate the Pearson correlation between the weekly numbers of cases from PU and those from nearby communities.
Confidence intervals are calculated using Fisher's Z transformation, and the corresponding significance is calculated using asymptomatic t approximation \citep{hollander2013nonparametric}.
The standard method of calculating Perason correlations and the associated significance assumes that the samples are independent and identically distributed---these assumptions generally do not hold for epidemic time series data, which are autocorrelated and non-stationary.
Nonetheless, the Pearson correlation has been the standard measure in epidemic time series analyses and can provide useful information about the strength of spatial coupling \citep{rohani1999opposite,grenfell2001travelling,viboud2006synchrony,baker2019epidemic}.
We note correlation analyses do not provide direct evidence for causality and therefore should be interpreted with care---we later investigate the role of community transmission using a mathematical model.''

\revtext{For example, it is stated that: "we are not able to infer the direction of causality—", but I am concerned this may be an understatement of the limitations with respect to making causal inference of any kind from the correlation results.}

Thank you for pointing this out.
We recognize that our phrasing was unclear. 
This sentence was meant to point at the limitation of our modeling analysis, rather than the correlation analysis.
We have tried to highlight the limitation of our correlation results clearer while making sure that we don't conflate conclusions from correlation analyses and modeling analyses:

``There are several limitations to our analysis.
While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, correlations do not provide causal link for community transmission.
For example, even if there was no mixing between campus and community populations, two populations can exhibit correlated epidemic dynamics if both populations are changing behavior in similar ways, reflecting changes in intervention measures in the greater New Jersey area.
However, it seems biologically implausible that there would be no mixing between the two populatins---other studies have also identified community transmission to be an important source of SARS-CoV-2 infections on campus \citep{fox2021response,hamer2021assessment}.
Our mathematical modeling illustrates that community transmission can explain the observed dynamics on PU campus, but we are not able to infer the direction of causality''

@@

\revtext{Reviewer \#2 Comments for the Author:}

\revtext{Thank you for the detailed point-by-point response from the authors. I think the revisions really help with the clarity of the manuscript and allow for a more reliable/realistic comparison between community and campus dynamics. I also appreciate the extended sensitivity analysis that accounts for some of the additional epidemiological changes during Omicron. I am happy to recommend this work for publication.}

### 2023-04-21 doc/university-covid.tex
campus done

Mathematical models have played a crucial role in exploring and guiding pandemic responses. 
University campuses present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
Here, we present descriptive and modeling analyses of SARS-CoV-2 transmission on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions and operational guidelines for the university campus.
We demonstrate strong spatiotemporal correlations in epidemic patterns between the university campus and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
Finally, we find that a super-spreading event likely played a major role in driving the Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.

@@

University campuses present challenges to preventing SARS-CoV-2 transmission, due to a high proportion of asymptomatic infections and high contact rates. SARS-CoV-2 outbreaks on the Princeton University campus offer an unusually well-documented perspective, rooted in mass asymptomatic testing, further informed by mathematical modeling aimed at guiding policy decisions. Here, we show that this model can parsimoniously capture observed outbreak patterns on campus during different eras of control. Our analysis reveals that strong coupling between epidemic dynamics on campus and in local communities drives the early epidemic. Subsequently, especially in the Omicron era, superspreading events came to dominate transmission on campus, thereby weakening the dynamical coupling of campus and community outbreaks.

@@

Across the first three semesters, we find strong and significant correlations between the weekly numbers of cases from PU and those from Mercer county:
fall 2020--2021 ($\rho = 0.79$ (95\% CI: 0.52--0.91; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.60--0.94; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.93$ (95\% CI: 0.84--0.97; $p < 0.001$), \fref{princeton}G). 
These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S3).
However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.47$ (95\% CI: $-0.18$--0.83; $p = 0.15$).
Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S3). 
We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S4)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases.
For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S4).

@@

Here, we analyze SARS-CoV-2 outbreaks on the PU campus between fall 2020 and early 2022.
We demonstrate strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
These correlations decreased with distance from Mercer County in fall 2021--2022, likely reflecting contact and commuting patterns as the university campus reopened.
Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of the 2020--2021 academic year and an increased frequency of infective community contacts during the fall semester of the 2021--2022 academic year, compared to previous semesters.
An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
Finally, our analysis highlights the potential role of super-spreading events in driving the spread of the Omicron variant on the PU campus.

@@

There are several limitations to our analysis.
While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
However, seeding from campus is unlikely: 
intervention measures on campus (e.g., frequent asymptomatic testing, contact tracing, and virtual classes during fall and spring semesters of 2020) likely limited onward transmission on campus.
In addition, even during periods of large Omicron outbreaks on campus in early 2022, the number of COVID-19 cases in Mercer County remained low, implying limited transmission from campus to community.
Decreasing patterns in epidemic correlations with distance further highlight the role of spatial spread in driving dynamics of SARS-CoV-2---such patterns are consistent with spatial spread of many other respiratory pathogens \citep{grenfell2001travelling, viboud2006synchrony, baker2019epidemic}.

@@

@@

\section*{Supplementary Figures}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[!htp]
\includegraphics[width=0.9\textwidth]{../figure_princeton_new/figure_princeton_testing.pdf}
\caption{
\textbf{The weekly number of PCR tests stratified by subpopulations.}
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000_bypop.pdf}
\caption{
\textbf{The weekly number of cases per 1000 stratified by subpopulations.}
The case trajectory for undergraduate students was omitted for the fall semester of 2020--2021 academic year due to a very low number of undergraduate students present on campus.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_correlation_per_1000.pdf}
\caption{
\textbf{Correlations between the weekly number of cases in PU and in Mercer County stratified by subpopulations.}
Points represent the number of reported cases.
Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
}
\end{figure}

\begin{figure}[!htp]
\begin{center}
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_map_new.pdf}
\end{center}
\caption{
\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey.}
(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey.
(D--F) Relationship between case correlations and distance from Mercer County.
Points represent the estimated correlation coefficients.
Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
Black points and lines represent correlations based on all cases in PU.
Red points and lines represent correlations based on cases among faculty and staff members in PU.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_newyork_new.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and New York City.}
(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
Red lines represent the weekly number of cases in New York City.
(D--F) Correlations between the weekly number of cases in PU and in New York City.
Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_philly_new.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and Philadelphia.}
(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
Red lines represent the weekly number of cases in Philadelphia.
(D--F) Correlations between the weekly number of cases in PU and in Philadelphia.
Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=0.8\textwidth]{../figure_princeton_new/figure_princeton_map_extended.pdf}
\caption{
\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.}
(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.
(D--F) Relationship between case correlations and distance from Mercer County.
Points represent the estimated correlation coefficients.
Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
Black points and lines represent correlations based on all cases in PU.
Red points and lines represent correlations based on cases among faculty and staff members in PU.
New York City is excluded from this analysis as the data provided by New York Times are not further stratified by county levels.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_diagram/diagram.pdf}
\caption{
\textbf{Compartmental diagram of the individual-based model.}
Each compartment represents a stage of infection: susceptible $S$, exposed $S$, pre-symptomatic $I_p$, symptomatic $I_s$, asymptomatic $I_a$, and recovered $R$.
Pre-symptomatic, symptomatic, and asymptomatic stages are further divided into two subcompartments.
Individuals in pre-symptomatic, symptomatic, and asymptomatic stages can test positive with 95\% sensitivity.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2020_all.pdf}
\caption{
\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2020.}
Points represent the weekly number of reported cases in PU.
Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
See figure 2 in the main text for details.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_spring_2020_all.pdf}
\caption{
\textbf{Comparisons between model predictions and the observed numbers of cases for spring 2020.}
Points represent the weekly number of reported cases in PU.
Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
See figure 2 in the main text for details.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2021_all.pdf}
\caption{
\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021.}
Points represent the weekly number of reported cases in PU.
Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
See figure 2 in the main text for details.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_nowaning.pdf}
\caption{
\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021 without immunity waning.}
Points represent the weekly number of reported cases in PU.
Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
See figure 2 in the main text for details.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
\caption{
\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a shorter latent and infectious periods.}
Solid lines represent median predictions.
Shaded areas represent 90\% quantiles across 100 simulations.
Points represent the observed data.
Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
Here, the assumed mean latent, presymptomatic, and asymptomatic stages of infection are 0.5 days shorter than the assumed values for simulations presented in Figure 3 in the main text.
}
\end{figure}

\begin{figure}[!htp]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
\caption{
\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a lower vaccine effectiveness.}
Solid lines represent median predictions.
Shaded areas represent 90\% quantiles across 100 simulations.
Points represent the observed data.
Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
Here, we assume 30\% vaccine effectiveness against infection.
}
\end{figure}

>>>>

Mathematical models have played a crucial role in exploring and guiding pandemic responses. 
University campuses present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
Here, we present descriptive and modeling analyses of SARS-CoV-2 transmission on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions and operational guidelines for the university campus.
Epidemic patterns between the university campus and surrounding communities exhibit strong spatiotemporal correlations.
Mathematical modeling analysis further suggests that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
Finally, we find that a super-spreading event likely played a major role in driving the Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.

@@

University campuses present challenges to preventing SARS-CoV-2 transmission, due to a high proportion of asymptomatic infections and high contact rates. SARS-CoV-2 outbreaks on the Princeton University campus offer an unusually well-documented perspective, rooted in mass asymptomatic testing, further informed by mathematical modeling aimed at guiding policy decisions. Here, we show that this model can parsimoniously capture observed outbreak patterns on campus during different eras of control. Our modeling analysis further reveals that strong coupling between epidemic dynamics on campus and in local communities drives the early epidemic. 
Subsequently, especially in the Omicron era, superspreading events came to dominate transmission on campus, thereby weakening the dynamical coupling of campus and community outbreaks.

@@

In order to compare the transmission dynamics on PU campus and in nearby communities, we calculate the Pearson correlation between the weekly numbers of cases from PU and those from nearby communities.
Confidence intervals are calculated using Fisher's Z transformation, and the corresponding significance is calculated using asymptomatic t approximation \citep{hollander2013nonparametric}.
The standard method of calculating Perason correlations and the associated significance assumes that the samples are independent and identically distributed---these assumptions generally do not hold for epidemic time series data, which are autocorrelated and non-stationary.
Nonetheless, the Pearson correlation has been the standard measure in epidemic time series analyses and can provide useful information about the strength of spatial coupling \citep{rohani1999opposite,grenfell2001travelling,viboud2006synchrony,baker2019epidemic}.
We note correlation analyses do not provide direct evidence for causality and therefore should be interpreted with care---we later investigate the role of community transmission using a mathematical model.

Across the first three semesters, we find strong and significant correlations between the weekly numbers of cases from PU and those from Mercer county:
fall 2020--2021 ($\rho = 0.79$ (95\% CI: 0.52--0.91; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.60--0.94; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.93$ (95\% CI: 0.84--0.97; $p < 0.001$), \fref{princeton}G). 
These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S3).
However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.47$ (95\% CI: $-0.18$--0.83; $p = 0.15$).
Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county (Supplementary Figure S3). 
We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S4)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases.
For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations (Supplementary Figure S4).

@@

Here, we analyze SARS-CoV-2 outbreaks on the PU campus between fall 2020 and early 2022.
We demonstrate strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
These correlations decreased with distance from Mercer County in fall 2021--2022, indicating weaker spatial coupling in the dynamics.
Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of the 2020--2021 academic year and an increased frequency of infective community contacts during the fall semester of the 2021--2022 academic year, compared to previous semesters.
An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
Finally, our analysis highlights the potential role of super-spreading events in driving the spread of the Omicron variant on the PU campus.

@@

There are several limitations to our analysis.
While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, correlations do not provide causal link for community transmission.
For example, even if there was no mixing between campus and community populations, two populations can exhibit correlated epidemic dynamics if both populations are changing behavior in similar ways, reflecting changes in intervention measures in the greater New Jersey area.
However, it seems biologically implausible that there would be no mixing between the two populatins---other studies have also identified community transmission to be an important source of SARS-CoV-2 infections on campus \citep{fox2021response,hamer2021assessment}.
Our mathematical modeling illustrates that community transmission can explain the observed dynamics on PU campus, but we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
However, seeding from campus is unlikely: 
intervention measures on campus (e.g., frequent asymptomatic testing, contact tracing, and virtual classes during fall and spring semesters of 2020) likely limited onward transmission on campus.
In addition, even during periods of large Omicron outbreaks on campus in early 2022, the number of COVID-19 cases in Mercer County remained low, implying limited transmission from campus to community.
Decreasing patterns in epidemic correlations with distance further highlight the role of spatial spread in driving dynamics of SARS-CoV-2---such patterns are consistent with spatial spread of many other respiratory pathogens \citep{grenfell2001travelling, viboud2006synchrony, baker2019epidemic}.

@@

\section*{Author contribution}

All authors contributed equally to the research design, data collection, formal analysis, interpretation of the results, and writing of the paper.

@@
